---
title: "Vg Analysis"
author: "Reed A. Cartwright"
format: pdf
editor: source
pdf-engine: lualatex
mainfont: TeX Gyre Pagella
monofont: Source Code Pro
geometry:
  - margin=0.5in
---

```{r}
#| include=FALSE

library(stringdist)
library(tidyverse)
library(vcfR)
library(here)
```

## Vizualizing Haplotypes

We have created a VCF of the Vg region based on 543 diploid samples, and for
this analysis we will investigate the biallelic sites with known ancestral
allele.

```{r}
raw_vcf <- read.vcfR(here("results", "apis_mellifera_vg.dna.vcf"),
                     verbose = FALSE)
raw_vcf_fix <- vcfR2tidy(raw_vcf, info_only = TRUE)$fix

vcf <- raw_vcf[is.polymorphic(raw_vcf) &
               is.biallelic(raw_vcf) & 
               raw_vcf_fix$AA != "."]
vcf_fix <- vcfR2tidy(vcf, info_only = TRUE)$fix
```

Next we will extract phased haplotytpes and encode variants as 0 = ancestral
and 1 = derived.

```{r}
haplotypes <- (extract.haps(vcf, verbose = FALSE) != vcf_fix$AA)*1L
num_sites <- nrow(haplotypes)
haplotypes_str <- apply(haplotypes, 2, str_flatten)
```

Next we will count how abundant each haplotype is in our dataset and plot the
haplotypes in two dimensions using principle-coordinate analysis (PCoA).

```{r}
tab <- tibble(hap = haplotypes_str) |> count(hap)

hap_dist <- stringdistmatrix(tab$hap, method = "hamming")
v <- cmdscale(hap_dist/nrow(haplotypes), 2)

tab <- tab |> mutate(x = v[,1], y = v[,2])
gg <- ggplot(tab, aes(x = x, y = y, size = n)) + geom_point(alpha = 0.5)
gg <- gg + xlab("Axis 1") + ylab("Axis 2")
gg <- gg + scale_size_continuous("Count")
gg <- gg + coord_fixed()
gg
```

The spread along Axis 1 reflects the presence of two haplogroups that are
present in our dataset at high frequency. When we look at a historgram of
MAF values, we see a many sites with a MAF \> 0.25.

```{r}
gg <- ggplot(vcf_fix, aes(x = MAF)) + geom_histogram(bins = 20)
gg
```

## Vizualizing Haplogroups

To look at the haplogroups, we will focus only on sites with a MAF \> 0.25.
Looking only at these sites, we determine that the two most common haplotypes,
do not share any derived alleles. This suggests that each haplogroup evolved
independent and in an isolated population. We will use these two haplotypes as
reference haplotypes for each haplogroup.

```{r}
# select sites with high MAF
hg_o <- vcf_fix$MAF >= 0.25
hg_pos <- which(hg_o)

# Customize AA based on recombination data
#vcf_fix$AA[hg_pos[47]] <- "C"

# create haplogroup sequences
haplotypes_hg <- (extract.haps(vcf[hg_o], verbose = FALSE) !=
                    vcf_fix[hg_o, ]$AA)*1L
haplotypes_hg_str <- apply(haplotypes_hg, 2, str_flatten)

out_hg <- vcf_fix[hg_o, ] |> select(REF, ALT, AA, Cerana:Florea) |>
  mutate(across(Cerana:Florea, \(x) case_when(
    x == "." ~ "?",
    x == AA ~ "0",
    x == REF ~ "1",
    x == ALT ~ "1",
    TRUE ~ "9"
  )))
out_hg <- list(
  cerana = str_flatten(out_hg$Cerana),
  dorsata = str_flatten(out_hg$Dorsata),
  laboriosa = str_flatten(out_hg$Laboriosa),
  florea = str_flatten(out_hg$Florea)
)

# Count haplogroup sequences
tab_hg <- tibble(hap = haplotypes_hg_str) |> count(pick(everything()))
tab_hg |> slice_max(n, n = 5)
```

If we plot the haplogroup-defining sites in two dimensions we see that the two
references separate on different ends of Axis 1.

```{r}
hap_dist_hg <- stringdistmatrix(tab_hg$hap, method = "hamming")
v <- cmdscale(hap_dist_hg/nrow(haplotypes_hg), 2)

tab_hg <- tab_hg |> mutate(x = v[,1], y = v[,2])
gg <- ggplot(tab_hg, aes(x = x, y = y, size = n)) + geom_point(alpha = 0.5)
gg <- gg + xlab("Axis 1") + ylab("Axis 2")
gg <- gg + scale_size_continuous("Count")
gg <- gg + coord_fixed()
gg
```

Now we will redraw the original plot, but colored according to haplogroup.

```{r}
hg_ref <- tab_hg |> slice_max(n, n = 1) |> pull(hap)

tab$hg <- tab$hap |> str_sub_all(which(hg_o), which(hg_o)) |>
    map_chr(str_flatten)
tab$hg_dist <- stringdistmatrix(tab$hg, hg_ref, "hamming")[,1]/sum(hg_o)

gg <- ggplot(tab, aes(x = x, y = y, size = n, color = hg_dist)) +
  geom_point()
gg <- gg + xlab("Axis 1") + ylab("Axis 2")
gg <- gg + scale_size_continuous("Count")
gg <- gg + scale_color_distiller("Haplogroup", palette = "RdYlBu")
gg <- gg + coord_fixed()
gg
```

## Recombinant Sequences

We can identify recombinant sequences by looking at sequences that have parts
of both haplogroups

```{r}
#| fig.height=5
hg_v <- str_split(c(tab$hg,out_hg), "")
hg_ref <- str_split_1(hg_ref, "")
hg_str <- map(hg_v, \(x) case_when(
    x == "0" ~ "0",
    x == hg_ref ~ "1",
    x == "1" ~ "2",
    x == "?" ~ "?",
    TRUE ~ "9"
    ))
hg_str <- map_chr(hg_str, str_flatten)

tab$hg_str <- hg_str[seq_len(nrow(tab))]

# Looking at singltons identifed as tripple recombinatns, we probably
# need to switch some ancestral alleles
xx <- str_count(tab$hg_str, "0*(?:1[01]*2[01]|2[02]*1?[02]*)")
yy <- str_count(old_hg_str, "0*(?:1[01]*|2[02]*)")
str_view(tab$hg_str[xx == 3], "0*(?:1[01]*(?:2[01]+)?|2[02]*(?:1[02]+)?)")

hg_str_v <- str_split(tab$hg_str, "")
n_rec <- map_int(hg_str_v, function(x) {
  # rle all sites that are not 0
  r <- rle(x[x != "0"])
  r$values[r$lengths == 1L] <- "0"
  x <- inverse.rle(r)
  r <- rle(x[x != "0"])
  
  length(r$lengths)
})

tab$is_recomb <- str_detect(tab$hg_str, "10*1") &
  str_detect(tab$hg_str, "20*2")

gg <- ggplot(tab, aes(x = x, y = y, size = n, color = hg_dist,
                      shape = is_recomb)) + geom_point()
gg <- gg + xlab("Axis 1") + ylab("Axis 2")
gg <- gg + scale_size_continuous("Count")
gg <- gg + scale_shape_discrete("Recombinant")
gg <- gg + scale_color_distiller("Background", palette = "RdYlBu")
gg <- gg + coord_fixed()
gg
```

Same analysis as before but now we will split it based on broad geographic regions

```{r}
#| fig.height=8
tab <- tibble(sample = colnames(haplotypes), hap = haplotypes_str)

tab <- tab |> separate_wider_regex(sample, c("[^_]+_",
    "apiary" = "..", ".*"), cols_remove = FALSE) |>
    separate_wider_delim(sample, "_", names = c("id", "id", NA), cols_remove = FALSE)
tab <- tab |> mutate(region = case_match(apiary,
    c("NC", "CA", "MD", "AZ", "MN", "IL") ~ "USA",
    c("FN", "SW", "RN", "CS", "DK", "IR", "PO", "TX", "FR") ~ "Brownbee",
    c("SL", "IT", "PL", "MK", "MT", "TR") ~ "South Europe"
    ))
tab <- tab |> count(hap, region)

hap_dist <- stringdistmatrix(tab$hap, method = "hamming")
v <- cmdscale(hap_dist/num_sites, 2)
tab <- tab |> mutate(x = v[,1], y = v[,2])
tab <- tab |> arrange(desc(n))

gg <- ggplot(tab, aes(x = x, y = y, size = n, color = region)) +
  geom_point(alpha = 0.5)
gg <- gg + xlab("Axis 1") + ylab("Axis 2")
gg <- gg + scale_size_continuous("Count")
gg <- gg + scale_color_brewer("Region", palette = "Set1")
gg <- gg + coord_fixed()
gg <- gg + facet_wrap(vars(region), ncol = 1)
gg

```
